Whole genome alignment with wavefront inception

### 2022-04-02 manuscript/fig_wflamba.tex
introduction
tweaks

\begin{figure}[ht!]
    \begin{subfigure}{\linewidth}
        %\caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/wflign.png}
        \label{fig:bad-sorting}
    \end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
    \caption{
      \textit{wflign} exploration of the alignment of $\sim$560kbp homologus segments of two \textit{Escherichia coli} genomes (GenBank NC\_000913.3 and NC\_008563.1).
      Gray indicates cells explored by WF$\lambda$.
      Blue cells are attempted, failed local WFA alignments, while green cells indicate alignments that match the maximum local alignment score.
      %The traceback (not shown) drives through the 
    }
    \label{fig:wflambda}
\end{figure}

>>>>

\begin{figure}[t!]
    \begin{subfigure}{\linewidth}
        %\caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/wflign.png}
        \label{fig:bad-sorting}
    \end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
    \caption{
      WFlign exploration of the alignment of $\sim$560kbp homologus segments of two \textit{Escherichia coli} genomes (GenBank NC\_000913.3 and NC\_008563.1).
      Gray indicates cells explored by WF$\lambda$.
      Blue cells are attempted, failed local WFA alignments, while green cells indicate alignments that match the maximum local alignment score.
      %The traceback (not shown) drives through the 
    }
    \label{fig:wflambda}
\end{figure}

### 2022-04-02 manuscript/main.tex
tweaks

\title[Whole-genome pairwise alignment]{Whole-genome pairwise alignment using the hierarchical wavefront algorithm}
\author[Guarracino \textit{et~al}.]{
Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}\text{\sfb 1}}$,
Njagi~Mwaniki\,$^{\text{\sfb 2}}$,
Santiago~ Marco-Sola\,$^{\orcidlink{0000-0001-7951-3914}\text{\sfb 3, 4}}$,
Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 5}*}$
}

@@

However, increasing sequence read lengths, and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary to achieve near-perfect alignment performance when mapping long reads to reference genomes.
Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \cite{Roberts_2004}.
Rather than considering all $k$-mers, \textit{windowed minimizer} sampling collects $\sim L/w$ $k-mers$ from a sequence of length $L$, wherein each $k$-mer minimizes a given hash function within a window of size $w$.
This sampling scheme provides guarantees about the distribution of sampled $k$-mers, which aids read alignment and mapping chaining.
However, it has been discovered that window minimziers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
In contrast, \textit{world minimizers}, which sparsify over the entire potential space of $k$-mers \citep{Broder_1997}, are not biased, and have proved useful as the basis for the whole genome comparison system Mash \citep{Ondov_2016}.

@@

@@

However, obtaining a base level alignment required a completely new approach to sequence alignment.
We exploit the $O(s^2)$ scaling properties of the WaveFront Algorithm (WFA) \citep{Marco_Sola_2020} in a new hierarchical arrangement (the ``we flying'' alignment algorithm, WFlign) wherein WFA is applied not to the matching of single characters of input genomes, but rather to the matching of segments of hundreds of base-pairs (e.g. 256bp) (WF$\lambda$).
These segments are matched by sketching their kmers, and where similarity is above a given threshold, an ``incepted'' WFA alignment is completed at the base level.
By taking half-overlapping segments, the method has a high probability of finding the optimal base-level alignment within the incepted WFA problems, and we ultimately combine these during traceback through the WF$\lambda$ matrix.
A final round of erosion and patching of the resulting alignment (with WFA) yields a base-level alignment.
For a segment length $L$, wavefront inception provides $O(4L^2)$ reduction in alignment complexity relative to direct application of WFA.
When $L=256$, this corresponds to a 16,384-fold reduction in computational complexity, a property born out by our experimental results.
Although WFlign is an approximate alignment algorithm not guaranteed to find the optimal alignment through the full matrix, here we show that our implementation nevertheless achieves state-of-the-art accuracy for variant detection via alignment.

>>>>

\title[Wavefront inception]{Whole genome alignment with wavefront inception}
\author[Guarracino \textit{et~al}.]{
Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}\text{\sfb 1}}$,
Njagi~Mwaniki\,$^{\text{\sfb 2}}$,
Santiago~ Marco-Sola\,$^{\orcidlink{0000-0001-7951-3914}\text{\sfb 3, 4}}$,
Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 5}*}$
}

@@

Increasing sequence read lengths and the arrival of whole genome assemblies for many species rendered such approaches more sensitive than necessary.
%The new problem of mapping long reads to reference genomes led to 
Methods like minimap2 \citep{Li_2018}, LRA \citep{Ren_2021}, Winnowmap \citep{Jain_2020}, and MashMap \citep{Jain_2018} compute mappings, or candidate alignments, using a sparse subset of all $k$-mers, which are often referred to as ``minimizers'' due to their definition as hash-minimizing $k$-mers in a given window \cite{Roberts_2004}.
Rather than considering all $k$-mers, \textit{windowed minimizer} sampling collects $\sim L/w$ $k-mers$ from a sequence of length $L$, wherein each $k$-mer minimizes a given hash function within a window of size $w$.
This sampling scheme provides guarantees about the distribution of sampled $k$-mers, which aids read alignment and mapping chaining.
It has been discovered that window minimziers are biased and may underestimate the true similarity of sequences \citep{Belbasi_2022}.
In contrast, \textit{world minimizers}, which sparsify over the entire potential space of $k$-mers \citep{Broder_1997}, are not biased, and have proved useful as the basis for the whole genome comparison system Mash \citep{Ondov_2016}.

@@

%Minimizer seeds have also proven useful for the generation of alignments.
%To obtain base level alignments using adaptive-banded Smith-Waterman-Gotoh requires seeds that initiate the correct alignment, rendering the use of minimizers necessary.
%But, this limits the level of further sparsification available to mapping methods.
%These costs are further exacerbated when obtaining the base level alignments.

\subsection{Current work}

@@

However, obtaining a base level alignment required a completely new approach to sequence alignment.
We exploit the $O(s^2)$ scaling properties of the WaveFront Algorithm (WFA) \citep{Marco_Sola_2020} in a new hierarchical arrangement (the ``we flying'' alignment algorithm, WFlign) wherein WFA is applied not to the matching of single characters of input genomes, but rather to the matching of segments of hundreds of base-pairs (e.g. 256bp) (WF$\lambda$).
These segments are matched by sketching their kmers, and where similarity is above a given threshold, an ``incepted'' WFA alignment is completed at the base level.
By taking half-overlapping segments, the method has a high probability of finding the optimal base-level alignment within the incepted WFA problems, and we ultimately combine these during traceback through the WF$\lambda$ matrix.
A final round of erosion and patching of the resulting alignment (with WFA) yields a base-level alignment.

For a segment length $L$, wavefront inception provides $O(4L^2)$ reduction in alignment complexity relative to direct application of WFA.
When $L=256$, this corresponds to a 16,384-fold reduction in computational complexity, a property born out by our experimental results.
Although WFlign is an approximate alignment algorithm not guaranteed to find the optimal alignment through the full matrix, here we show that our implementation nevertheless achieves state-of-the-art accuracy for variant detection via alignment.
